Toward nonlinear multigrid \\ for nonlinear variational inequalities

### 2023-02-07 talk/slides.tex
progress talk, plus reference structure
progress ... now dinner

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{empheq,bm,xspace,fancyvrb,soul}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,decorations.markings,decorations.pathreplacing,fadings,positioning}
\usepackage[kw]{pseudo}
\pseudoset{left-margin=15mm,topsep=5mm,idfont=\texttt,st-left=,st-right=}

@@

\newcommand{\mR}{R^{\bm{\oplus}}}
\newcommand{\iR}{R^{\bullet}}

@@

@@

\includegraphics[width=0.6\textwidth]{images/obstacle65.pdf} \qquad \includegraphics[width=0.3\textwidth]{images/obstacle-sets.png}

\bigskip
\only<1>{
\begin{itemize}
\item \emph{problem.} on a domain $\Omega \subset \RR^2$, find the displacement $u(x)$ of a membrane, with fixed value $u = g$ on $\partial \Omega$, above an \emph{obstacle} $\psi(x)$, which minimizes the elastic energy
    $$J(u) = \int_\Omega \frac{1}{2} |\grad u|^2 - f\, u$$
\item shown above: $\Omega=(-2,2)^2$, $\psi(x)$ a hemisphere
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item i.e.~constrained optimization over an \emph{admissible set}
	$$\mathcal{K} = \left\{v \in H^1(\Omega) \,:\, v\big|_{\partial \Omega} = g \text{ and } v \ge \psi\right\}$$
\item informally, $J'(u)$ points directly into $\mathcal{K}$ ({\color{FireBrick} \emph{variational inequality}; VI}):
    $$\ip{J'(u)}{v-u} = \int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \text{for all } v \in \mathcal{K}$$
\end{itemize}
}
\only<3>{
\begin{itemize}
\item the solution defines \emph{active} $A_u = \{u = \psi\}$ and \emph{inactive} $R_u = \{u> \psi\}$ subsets of $\Omega$, and a \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
\item a naive strong form poses the problem in terms of its solution:
{\small
\begin{align*}
-\grad^2 u &= f \text{ on $R_u$} \\
u &= \psi \text{ on $A_u$}
\end{align*}
}
\item by contrast, the weak form VI is meaningful

\phantom{x}
\end{itemize}
}
\only<4>{
\begin{itemize}
\item a \emph{complementarity problem} (CP) is a meaningful strong form:
{\small
\begin{align*}
u - \psi &\ge 0 \\
-\grad^2 u - f &\ge 0 \\
(u - \psi)(-\grad^2 u - f) &= 0
\end{align*}
}
\item for optimization problems: CP $=$ KKT conditions

@@

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous, generally nonlinear operator
    \begin{itemize}
    \item[$\circ$] $F$ may be defined only on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not} be the derivative of a $J(u)$
    \end{itemize}
\item the general problem {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\end{itemize}
\end{frame}

@@

\vspace{-2mm}
{\color{FireBrick} VI($F$,$\mathcal{K}$)} for $u \in \mathcal{K}$:
$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}}$$
\end{minipage}
\end{tabular}
\end{center}
\end{frame}

@@

\begin{itemize}
\item elastic contact, Signorini problems (e.g.~Kikuchi \& Oden 1988)
\item viscous contact problems (de Diego et al.~2022)
\item pricing of American options in the Black-Scholes model
\item the geometry of glaciers \dots \emph{more soon}

@@

\begin{frame}{a nonlinear 2-grid scheme}

\begin{center}
\includegraphics[width=0.3\textwidth]{images/two-grids.png}
\end{center}

\bigskip
\only<1>{
\begin{itemize}
\item suppose we have a nonlinear elliptic PDE problem:
	$$F(u) = 0$$
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item x
\end{itemize}
}
\end{frame}

\begin{frame}{nonlinear multigrid}

\mbox{
\includegraphics[height=0.10\textheight]{images/mg-grids.png} \qquad \includegraphics[height=0.15\textheight]{images/mg-vcycle.png} \quad
\includegraphics[height=0.15\textheight]{images/mg-fcycle.png}
}

\bigskip
\only<1>{
\begin{itemize}
\item suppose we have a nonlinear elliptic PDE problem:
	$$F(u) = 0$$
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item x
\end{itemize}
}
\end{frame}

@@

>>>>

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{empheq,bm,xspace,fancyvrb,soul}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,decorations.markings,decorations.pathreplacing,fadings,positioning}
\usepackage[kw]{pseudo}
\pseudoset{left-margin=15mm,topsep=5mm,label=,idfont=\texttt,st-left=,st-right=}

@@

@@

\newcommand{\maxR}{R^{\bm{\oplus}}}
\newcommand{\minR}{R^{\bm{\ominus}}}
\newcommand{\iR}{R^{\bullet}}

@@

\includegraphics[width=0.55\textwidth]{images/obstacle65.pdf} \qquad \includegraphics[width=0.35\textwidth]{images/obstacle-sets.png}

\bigskip
\only<1>{
\begin{itemize}
\item \emph{problem.} on a domain $\Omega \subset \RR^2$, find the displacement $u(x)$ of a membrane, with fixed value $u = g$ on $\partial \Omega$, above an \emph{obstacle} $\psi(x)$, which minimizes the elastic energy
    $$J(v) = \int_\Omega \frac{1}{2} |\grad v|^2 - f\, v$$
\item shown above: \quad $\Omega=(-2,2)^2$, $\psi(x)$ a hemisphere, $f(x)=0$
\end{itemize}

\phantom{x}
}
\only<2>{
\begin{itemize}
\item i.e.~constrained optimization over an \emph{admissible set}
	$$\mathcal{K} = \left\{v \in H^1(\Omega) \,:\, v\big|_{\partial \Omega} = g \text{ and } v \ge \psi\right\}$$
\item $J'(u)$ points directly into $\mathcal{K}$, the {\color{FireBrick} \emph{variational inequality} (VI)}:
    $$\ip{J'(u)}{v-u} = \int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \text{for all } v \in \mathcal{K}$$
\end{itemize}
}
\only<3>{
\begin{itemize}
\item the solution defines \emph{active} $A_u = \{u = \psi\}$ and \emph{inactive} $R_u = \{u> \psi\}$ subsets of $\Omega$, and a \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
\item naive strong form would pose the problem in terms of its solution:
\begin{align*}
-\grad^2 u &= f \quad \text{ on $R_u$} \\
u &= \psi \quad \text{ on $A_u$}
\end{align*}

\phantom{x}
\end{itemize}
}
\only<4>{
\begin{itemize}
\item the \emph{complementarity problem} (CP) is meaningful as a strong form:
\begin{align*}
u - \psi &\ge 0 \\
-\grad^2 u - f &\ge 0 \\
(u - \psi)(-\grad^2 u - f) &= 0
\end{align*}
   \begin{itemize}
   \item[$\circ$] for optimization problems: \quad CP $=$ KKT conditions
   \end{itemize}

@@

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous, generally nonlinear operator
    \begin{itemize}
    \item[$\circ$] $F$ may be defined only on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not}\, be the derivative of an objective function $J$
    \end{itemize}
\item the general problem {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\end{itemize}
\end{frame}

@@

\vspace{-2mm}
{\color{FireBrick} VI} for $u \in \mathcal{K}$:
$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}}$$
\end{minipage}
\end{tabular}
\end{center}
\end{frame}

@@

\begin{itemize}
\item elastic contact, Signorini problems (e.g.~Kikuchi \& Oden 1988)
\item viscous contact problems (de Diego et al.~2022)
\item pricing of American options in the Black-Scholes model
\item the geometry of glaciers \hfill $\longleftarrow$ \emph{more soon}

@@

\begin{frame}{nonlinear 2-grid scheme}

\begin{center}
$\Omega^h$\, \includegraphics[height=0.16\textheight]{images/fine-grid.png} \hspace{25mm} \includegraphics[height=0.16\textheight]{images/coarse-grid.png} \,$\Omega^H$
\end{center}

\only<1>{
\begin{itemize}
\item consider a nonlinear elliptic PDE problem:
	$$F(u) = \ell$$

	\begin{itemize}
	\item for example, $F : \mathcal{V} \to \mathcal{V}'$ for $\mathcal{V}=H^1(\Omega)$, with $\ell\in \mathcal{V}'$
	\end{itemize}
\item FEM/FD discretization gives algebraic system on fine grid $\Omega^h$:
    $$F^h(u^h) = \ell^h$$
\item suppose $w^h$ yields residual norm $\|\ell^h - F^h(w^h)\| > \text{TOL}$
\end{itemize}

\phantom{x}
}
\only<2-3>{
\begin{itemize}
\item how can we improve $w^h$ \emph{without} globally linearizing $F^h$?

are there alternatives to Newton's method?
\item note the \emph{residual} $r^h(w^h) = \ell^h - F^h(w^h)$ is computable,

while the \emph{error} $e^h = w^h-u^h$ is unknown
\item the residual definition can be rewritten
    $$F^h(w^h) - F^h(u^h) \stackrel{\ast}{=} -r^h(w^h)$$
\only<2>{\item for $F^h$ linear, try to solve this \emph{error eqn} $F^h(e^h) = -r^h(w^h)$ for $\tilde e^h$, and correct $w^h \leftarrow w^h-\tilde e^h$ to improve $w^h$?}
\only<3>{\item \st{for $F^h$ linear, try to solve this \emph{error eqn} $F^h(e^h) = -r^h(w^h)$ for $\tilde e^h$, and correct $w^h \leftarrow w^h-\tilde e^h$ to improve $w^h$?}}
\end{itemize}
}
\only<4>{
\begin{itemize}
\item \emph{nodewise problem}: for $\psi_j^h$ a hat function or dof, solve for $c\in\RR$:
	$${\color{FireBrick} \phi_j(c) = r^h(w^h + c \psi_j^h)[\psi_j^h] = 0}$$
\item sweeping through and solving nodewise problems is a \emph{smoother}
    \begin{itemize}
    \item[$\circ$] Fourier analysis on linear PDEs shows smoothing property
    \item[$\circ$] post-smoothing, $w^h-u^h$ and $r^h(w^h)$ have small high-frequency
    \end{itemize}
\item post-smoothing, both sides of $\ast$ should be accurately approximate-able on a coarser grid
\end{itemize}
}
\only<5>{
\begin{itemize}
\item Brandt (1977) proposed to replace $F^h(w^h) - F^h(u^h) = -r^h(w^h)$ with a coarse-grid, re-discretized version
\item \emph{full approximation storage} (FAS) \emph{equation}:
	$${\color{FireBrick} F^H(w^H) - F^H(\iR w^h) = -R r^h(w^h)}$$

    \begin{itemize}
    \item[$\circ$] $\iR:\mathcal{V}^h \to \mathcal{V}^H$ is \emph{injection}
    \item[$\circ$] $R:(\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
    \item[$\circ$] if $w^h=u^h$ exactly then $w^H = \iR w^h$
    \end{itemize}

FIXME CHECK CAREFULLY FROM HERE
\item equivalently ${\color{FireBrick} F^H(w^H) = \ell^H}$ where ${\color{FireBrick} \ell^H = FIXME R\,F^h(u^h) - \iR\, r^h(w^h)}$
\end{itemize}
}
\end{frame}

\begin{frame}{FAS 2-grid solver}

\begin{align*}
&\text{smooth by sweeps over $j$:}  & &w^h \leftarrow \left[\phi_j(c) = 0 \,\forall j\right] \\
&\text{evaluate:}                   & &\ell^H = R\,F^h(u^h) - \iR\, r^h(w^h) \\
&\text{solve:}                       & &F^H(w^H) = \ell^H \\
&\text{correct}:                    & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{smooth by sweeps over $j$:}  & &w^h \leftarrow \left[\phi_j(c) = 0 \,\forall j\right]
\end{align*}

\bigskip
\begin{itemize}
\item evaluate$+$solve$+$correct \, $=$ \, \emph{coarse grid correction}
\end{itemize}
\end{frame}

\begin{frame}{nonlinear multigrid = FAS V-cycle or F-cycle}

\includegraphics[height=0.15\textheight]{images/mg-grids.png}

\bigskip

\hfill \mbox{
\includegraphics[height=0.15\textheight]{images/mg-vcycle.png} \quad
\includegraphics[height=0.15\textheight]{images/mg-fcycle.png}
}

\vspace{-3mm}

\only<1>{
{\small
\begin{pseudo}
\pr{fas-vcycle}: \\+
    for $j=J$ downto $j=1$ \\+
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j;u^j)$ \\
      $u^{j-1} = \iR u^j$ \\
      $\ell^{j-1} = f^{j-1}(u^{j-1}) + R \left(\ell^j - f^j(u^j)\right)$ \\-
    $\text{\pr{solve}}(\ell^0;u^0)$ \\
    for $j=1$ to $j=J$ \\+
      $u^j \gets u^j + P (u^{j-1} - \iR u^j)$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j;u^j)$ \\-
\end{pseudo}
}
}
\only<2>{
\begin{itemize}
\item x
\end{itemize}
}
\end{frame}

@@

\begin{frame}{\emph{background} references}

{\scriptsize
%{\notsotiny
\input{background.tex}
}
\end{frame}
